---
title: "index"
author: "Betsy"
date: "November 30, 2017"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true 


---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## `Scatterplot`

```{r mpg.scatter}


#data()
#?mpg
#mpg

g <- ggplot(data = mpg, aes(x = displ, y = hwy))
g + geom_point()


```


### I need `colour!`

```{r colour}

g + geom_point(aes(color = class))

```

### Funky `shape` time -- variation in car class in black & white 

```{r b&w}

ggplot(data = mpg, aes(x=displ, y = hwy, shape = class)) + 
  geom_point()

```

#### but shapes max out at six types so be careful....



### `shape size`: need a figure in black & white
```{r shape size}

ggplot(data = mpg, aes(x = cty, y = hwy, size = class)) + geom_point()

```

## Car Class determined by `size` and Fuel Type determined by `colour`
```{r colour and size}

ggplot(data = mpg, aes(x = cty, y = hwy, size = class, color = fl)) + 
  geom_point(alpha = 0.2)

```



```{r displ feature}

ggplot(data = mpg) + 
  geom_point(mapping = aes(colour = displ < 5, x = displ, y = hwy))
   

```


## Multiple geoms 

```{r mult_geoms}

ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point() + 
  geom_smooth()


```


## Customize

```{r customize}

ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point() + 
  geom_smooth() + 
  labs(title = "Relationship between engine size and MPG", 
       x = "Highway MPG",
       y = "Engine displacement (litters)") + 
  theme_classic()


```

## Challenge 
```{r challenge}

ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = hwy , size = class)) + 
  geom_smooth(method = 'lm') + 
  labs(title = "Relationship between engine size and MPG", 
       x = "Highway MPG",
       y = "Engine displacement (litters)") + 
  theme_classic()+
  theme(text = element_text(size = 16))


```





## Data Wrangling with `dplyr`

```{r clean up}

gapminder <- read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder.csv")

gapminder

# useful functions to get to know your data:

head(gapminder) #will show first 6 rows of the data 
tail(gapminder) # show last 6 rows of the data

head(gapminder, 10) #can tell the program how many rows to show 
tail(gapminder, 9)

#looking at the structure of your data, the column has what type of information, like chr = character, int = interger, num = numerical values
str(gapminder)

#return the column names
names(gapminder)

#dimensions of the data set 
dim(gapminder)
ncol(gapminder)
nrow(gapminder)

#build the dimension output from ncol and nrow --> combine rows and columns 
c(nrow(gapminder), ncol(gapminder))

#summary statistics
summary(gapminder)

#everything above we were operating on the whole gapminder data set 

head(gapminder$lifeExp)

```

`read_csv` is from the `readr` package (part of the tidyverse). It is not part of `read.csv`, which is part of base R. 

#Practice with Verbs 

### Filter 
```{r filter}

swede <- filter(gapminder, country == "Sweden")
swede_mean <- mean(swede$lifeExp)
mean(filter(gapminder, country == "Sweden")$lifeExp)

# pipe operator `%>%`

##same output
gapminder %>% head(3)
head(gapminder,3)

##let's use the %>% with our Sweden example
swede <- filter(gapminder, country == "Sweden")
swede_mean <- mean(swede$lifeExp)

gapminder %>% filter(country == "Sweden") %>% summarise(mean(lifeExp))

```



###Select

```{r select}

#select 2 columns 
select(gapminder, year, lifeExp)
gapminder %>% filter(country == "Sweden") %>% select(year, lifeExp)

gapminder %>% 
  filter(country == "Cambodia") %>% 
  select(-continent, -lifeExp)

gapminder %>% 
  filter(country =="Mexico", year == 2002) %>% 
  select(-continent)

```

###Mutate
```{r mutate}

#create a new column with a function of two other columns
GDPdata <- gapminder %>% 
  mutate(gdp = pop * gdpPercap)

#let's add an index
GDPdata %>% 
  mutate(index = 1:nrow(gapminder)) %>% 
  select(-continent, -lifeExp)
 
gapminder %>% 
  mutate(gdp = pop * gdpPercap, 
        index = 1:nrow(gapminder)) %>% 
  select(-continent)

```

###Challenge in verbs 
```{r challenge2}

#find the maximum gdpPercap for Egypt and Vietnam, create a new column 
# %in% operator lets you filter multiple things within a `c()`

gapminder %>% 
  filter(country %in% c("Egypt", "Vietnam")) %>% 
  mutate(max_gdpPercap = max(gdpPercap))

#group_by() so that we can get2 maxes
gapminder %>% 
  filter(country %in% c("Egypt", "Vietnam")) %>% 
  group_by(country) %>% 
  summarise(max_gdpPercap = max(gdpPercap))


# let's leep hte year associated with that max_gdpPercap

gapminder %>% 
  group_by(country) %>% 
  mutate(max_gdpPercap = max(gdpPercap)) %>% 
  filter(max_gdpPercap == gdpPercap) %>% 
  arrange(desc(max_gdpPercap))

#if you wanted descending order, arrange(desc(max_gdpPercap))
```


#Day 2

## joining datasets

```{r co2 emissions}

# read in co2 emissions 
co2 <- read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/co2.csv")

co2 %>% head()
co2 %>% str

#create new variable with only gapminder 2007 data 
gap_2007 <- gapminder %>% 
  filter(year == 2007)
gap_2007 %>% head()
gap_2007 %>% str

#left_join gap_2007 to co2
lj <- left_join(gap_2007, co2, by = "country")
lj %>% dim() #124 rows

#right_join gap_2007 to co2
rj <- right_join(gap_2007, co2)
rj %>% dim()
rj %>% View()
```

## Reshaping datasets with `gather`, `spread`, `separate`, and `unite`

```{r reshaping datasets}
#wide format
gap_wide <- readr::read_csv('https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder_wide.csv')

gapminder <- readr::read_csv('https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder.csv')

head(gap_wide)
str(gap_wide)

```

###Gather

```{r}
#Use gather to turn `gap_wide` into a long format dataset

gap_long <- gap_wide %>% 
  gather(key = obstype_year, 
         value = obs_values,
         dplyr::starts_with("pop"),
         dplyr::starts_with("lifeExp"),
         dplyr::starts_with("gdpPercap"))

#When you have a big data set you can subtract the columns rather than including all of the ones you want 
gap_long <- gap_wide %>% 
  gather(key = obstype_year, 
         value = obs_values,
         - continent, -country)
gap_long

```

### Separate
```{r}
gap_long <- gap_wide %>% 
  gather(key = obstype_year, 
         value = obs_values,
         - continent, -country) %>% 
  separate(obstype_year,
           into = c("obs_type", "year"),
           sep = "_",
           convert = T)
head(gap_long)
tail(gap_long)

```

### Plot long format data 
```{r}
life_df <- gap_long %>% 
  filter(obs_type == "lifeExp", 
         continent == "Americas")
head(life_df)

ggplot(data = life_df, aes(x = year, y = obs_values, color = country)) + 
  geom_line()
```


### Using gap_long, calculate and plot the mean life expectancy for each continent over time from 1982 to 2007. Give your plot a title and assign x and y labels. 

```{r}
gap_life <- gap_long %>% 
  group_by(continent, obs_type, year) %>% 
  filter(obs_type == "lifeExp", 
         year >= 1982)
gap_life

gap_mean <- gap_life %>% 
  summarize(means = mean(obs_values)) 

gap_mean

ggplot(data = gap_mean, aes(x = year, y = means, color = continent)) + 
  geom_line()+
  labs(title = "Mean life expectancy for each continent from 1982 to 2007", 
       x = "Year",
       y = "Average Life Expectancy (years)")

```

### spread()

```{r}
head(gap_long)
gap_normal <- gap_long %>% 
  spread(key = obs_type, 
         value = obs_values)
gap_normal
```





