---
title: "index"
author: "Betsy"
date: "November 30, 2017"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true 


---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## `Scatterplot`

```{r mpg.scatter}


#data()
#?mpg
#mpg

g <- ggplot(data = mpg, aes(x = displ, y = hwy))
g + geom_point()


```


### I need `colour!`

```{r colour}

g + geom_point(aes(color = class))

```

### Funky `shape` time -- variation in car class in black & white 

```{r b&w}

ggplot(data = mpg, aes(x=displ, y = hwy, shape = class)) + 
  geom_point()

```

#### but shapes max out at six types so be careful....



### `shape size`: need a figure in black & white
```{r shape size}

ggplot(data = mpg, aes(x = cty, y = hwy, size = class)) + geom_point()

```

## Car Class determined by `size` and Fuel Type determined by `colour`
```{r colour and size}

ggplot(data = mpg, aes(x = cty, y = hwy, size = class, color = fl)) + 
  geom_point(alpha = 0.2)

```



```{r displ feature}

ggplot(data = mpg) + 
  geom_point(mapping = aes(colour = displ < 5, x = displ, y = hwy))
   

```


## Multiple geoms 

```{r mult_geoms}

ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point() + 
  geom_smooth()


```


## Customize

```{r customize}

ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point() + 
  geom_smooth() + 
  labs(title = "Relationship between engine size and MPG", 
       x = "Highway MPG",
       y = "Engine displacement (litters)") + 
  theme_classic()


```

## Challenge 
```{r challenge}

ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = hwy , size = class)) + 
  geom_smooth(method = 'lm') + 
  labs(title = "Relationship between engine size and MPG", 
       x = "Highway MPG",
       y = "Engine displacement (litters)") + 
  theme_classic()+
  theme(text = element_text(size = 16))


```





## Data Wrangling with `dplyr`

```{r clean up}

gapminder <- read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder.csv")

gapminder

# useful functions to get to know your data:

head(gapminder) #will show first 6 rows of the data 
tail(gapminder) # show last 6 rows of the data

head(gapminder, 10) #can tell the program how many rows to show 
tail(gapminder, 9)

#looking at the structure of your data, the column has what type of information, like chr = character, int = interger, num = numerical values
str(gapminder)

#return the column names
names(gapminder)

#dimensions of the data set 
dim(gapminder)
ncol(gapminder)
nrow(gapminder)

#build the dimension output from ncol and nrow --> combine rows and columns 
c(nrow(gapminder), ncol(gapminder))

#summary statistics
summary(gapminder)

#everything above we were operating on the whole gapminder data set 

head(gapminder$lifeExp)

```

`read_csv` is from the `readr` package (part of the tidyverse). It is not part of `read.csv`, which is part of base R. 

#Practice with Verbs 

### Filter 
```{r filter}

swede <- filter(gapminder, country == "Sweden")
swede_mean <- mean(swede$lifeExp)
mean(filter(gapminder, country == "Sweden")$lifeExp)

# pipe operator `%>%`

##same output
gapminder %>% head(3)
head(gapminder,3)

##let's use the %>% with our Sweden example
swede <- filter(gapminder, country == "Sweden")
swede_mean <- mean(swede$lifeExp)

gapminder %>% filter(country == "Sweden") %>% summarise(mean(lifeExp))

```



###Select

```{r select}


```